API design for machine learning software: experiences from the scikit-learn project

### 2013-09-09 paper.tex
trim/copyedit

\sklearn is an increasingly popular machine learning
library. Written in Python, it is designed to be simple and efficient, accessible to
non-experts, and reusable in various contexts. In this paper, we present and
discuss our design choices for the application programming interface (API) of
the project. In particular, we describe the simple and elegant interface shared
by all learning and processing units in the library and then discuss its
advantages in terms of composition and reusability. The paper also comments on
implementation details specific to the Python ecosystem and analyzes obstacles
faced by users and developers of the library.

@@

The \sklearn project\footnote{\url{http://scikit-learn.org}}
\citep{pedregosa2011} provides an open source machine learning
library for the Python programming language. The ambition of the project is to
provide efficient and well-established machine learning tools within a
programming environment that is accessible to non-machine learning experts and
reusable in various scientific areas. The project is not a novel domain-specific
language, but a library that provides machine learning idioms to a
general-purpose high-level language. Among other things, it includes classical learning
algorithms, model evaluation and selection tools, as well as
preprocessing procedures. The library is
distributed under the simplified BSD license, encouraging its use in both
academic and commercial settings.

@@

In \sklearn, we chose a representation of data that is as close as
possible to the matrix representation: datasets are encoded as NumPy
multidimensional arrays for dense data and as SciPy sparse matrices for sparse
data. While these may seem rather unsophisticated data representations when
compared to more object-oriented constructs, such as the ones used by
Weka \citep{hall2009weka}, they bring the prime advantage of allowing us to rely
on efficient NumPy and SciPy vectorized operations while keeping
the code short and readable.  This design choice has also been motivated by
the fact that, given their pervasiveness in many other scientific Python
packages, many scientific users of Python are already familiar with NumPy dense
arrays and SciPy sparse matrices.
From a practical point of view, these formats also provide a collection of
data loading and conversion tools which make them very easy to use in many
contexts. Moreover, for tasks where the inputs are text files or semi-structured
objects, we provide \textit{vectorizer} objects that efficiently convert such
data to the NumPy or SciPy formats.

For efficiency reasons, the public interface is oriented towards processing
batches of samples rather than single samples per API call. While classification
and regression algorithms can indeed make predictions for single samples,
\sklearn objects are not optimized for this use case. (The few online learning
algorithms implemented are intended to take mini-batches.) Batch processing makes
optimal use of NumPy and SciPy by preventing the overhead inherent to Python
function calls or due to per-element dynamic type checking. Although this might
seem to be an artifact of the Python language, and therefore an implementation
detail that leaks into the API, we argue that APIs should be designed so as not
to tie a library to a suboptimal implementation strategy. As such, batch
processing enables fast implementations in lower-level languages (where memory
hierarchy effects and the possibility of internal parallelization come into
play).

@@

From the start, the choice to let a single object serve dual purpose as
estimator and model has mostly been driven by usability and technical
considerations. From the user point of view, having two coupled instances (i.e.,
an estimator object, used as a factory, and a model object, produced by the
estimator) indeed decreases the ease of use and is also more likely to
unnecessarily confuse newcomers. From the developer point of view, decoupling
estimators from models also creates parallel class hierarchies and increases the
overall maintenance complexity of the project. For these practical reasons, we
believe that decoupling estimators from models is not worth the effort. A good
reason for decoupling however, would be that it makes it possible to ship a
model in a new environment without having to deal with potentially complex
software dependencies. Such a feature could however still be implemented in
\sklearn by making estimators able to export a fitted model, using the
information from its public attributes, to an agnostic model description such as
PMML~\citep{guazzelli2009pmml}.

@@

scaler = StandardScaler()
scaler.fit(X_train)
X_train = scaler.transform(X_train)
\end{pythoncode}
Of course, in practice, it is important to apply the same preprocessing to the
test data \texttt{X\_test}. Since a \texttt{StandardScaler} estimator stores the
mean and standard deviation that it computed for the training set, transforming
an unseen test set \texttt{X\_test} maps it into the appropriate region of
feature space:
\begin{pythoncode}
X_test = scaler.transform(X_test)
\end{pythoncode}
Transformers also include a variety of learning algorithms, such as
dimension reduction (PCA, manifold learning), kernel approximation,
and other mappings from one feature space to another.

Additionally, by leveraging the fact that \texttt{fit} always returns the
estimator it was called on, the \texttt{StandardScaler} example above can be
rewritten in a single line using method chaining:
\begin{pythoncode}
X_train = StandardScaler().fit(X_train).transform(X_train)
\end{pythoncode}

Furthermore, every transformer allows \texttt{fit(X\_train).transform(X\_train)}
to be written as \texttt{fit\_transform(X\_train)}.
The combined \texttt{fit\_transform} method prevents repeated computations.
Depending on the transformer,
it may skip only an input validation step,
or in fact use a more efficient algorithm for the transformation.
In the same spirit, clustering algorithms provide a
\texttt{fit\_predict} method
that is equivalent to \texttt{fit} followed by \texttt{predict},
returning cluster labels assigned to the training samples.

@@

Optionally, the model selection algorithms
also take a cross-validation scheme and a score function.  \sklearn provides
various such cross-validation schemes, including $k$-fold (default),
stratified $k$-fold and leave-one-out.
The score function used by default is the estimator's \texttt{score} method,
but the library provides a variety of
alternatives that the user can choose from,
including accuracy, AUC and $F_1$ score for classification,
$R^2$ score and mean squared error for regression.

@@

In other places of the library, in particular in the vectorization code
for unstructured input, the toolkit is also designed to be
extensible. Here, estimators provide hooks for user-defined code: objects or
functions that follow a specific API can be given as arguments at vectorizer
construction time. The library then calls into this code, communicating with it by passing objects of standard Python/NumPy types.
Again, such external user code can be kept agnostic of the \sklearn
class hierarchy.

Our rule of thumb is that user code should not be tied to \sklearn---which is a
\textit{library}, and not a \textit{framework}. This principle indeed avoids a
well-known problem with object-oriented design, which is that users wanting a
``banana'' should not get ``a gorilla holding the banana and the entire jungle''
(J.~Armstrong, cited by \citealp[p.~213]{seibel2009coders}).
That is, programs using \sklearn should not be intimately tied to it,
so that their code can be reused with other toolkits or in other contexts.

@@

Our implementation guidelines emphasize writing efficient but readable code.
In particular, we focus on making the codebase easily maintainable and
understandable in order to favor external contributions. Whenever practicable,
algorithms implemented in \sklearn are written in Python,
using NumPy vector operations for numerical work.
This allows for the code to remain concise, readable and
efficient. For critical algorithms that cannot be easily and efficiently
expressed as NumPy operations, we rely on Cython \citep{behnel2011cython}
to achieve competitive performance and scalability. Cython is a
compiled programming language that extends Python with static typing. It
produces efficient C extension modules that are directly importable from the
Python run-time system. Examples of algorithms written in Cython include
stochastic gradient descent for linear models, some graph-based clustering
algorithms and decision trees.

To facilitate the installation and thus adoption of \sklearn,
the set of external dependencies is kept to a bare minimum:
only Python, NumPy and SciPy are required for a functioning installation.
Binary distributions of these are available for the major platforms.
Visualization functionality depends on Matplotlib \citep{hunter2007matplotlib}
and/or Graphviz \citep{gansner2000},
but neither is required to perform machine learning or prediction.
When feasible, external libraries are integrated into the codebase.
In particular, \sklearn includes modified versions of \textsf{LIBSVM} and \textsf{LIBLINEAR}
\citep{chang2011libsvm,fan2008}, both written in C{}\verb!++!
and wrapped using Cython modules.

@@

Recent years have witnessed a rising interest in machine learning and data
mining with applications in many fields. With this rise comes a host of machine
learning packages (both open source and proprietary) with which \sklearn
competes. Some of those, including Weka~\citep{hall2009weka} or
Orange~\citep{Demsar2004}, offer APIs but actually focus on the use of a graphical user interface (GUI)
which allows novices to easily apply machine learning algorithms. By
contrast, the target audience of \sklearn is capable of programming, and
therefore we focus on developing a usable and consistent API, rather than expend
effort into creating a GUI\@. In addition, while GUIs are useful tools, they
sometimes make reproducibility difficult in the case of complex workflows
(although those packages usually have developed a GUI for managing complex
tasks).

Other existing machine learning packages
such as SofiaML\footnote{\url{https://code.google.com/p/sofia-ml}}~\citep{sculley2009large}
and Vowpal~Wabbit\footnote{\url{http://hunch.net/$\sim$vw}}
are intended to be used as command-line tools
(and sometimes do not offer any type of API).
While these packages have the advantage
that their users are not tied to a particular programming language,
the users will find that they still need programming to process input/output,
and will do so in a variety of languages.
By contrast, \sklearn allows users to implement that entire workflow
in a single program, written in a single language,
and developed in a single working environment.
This also makes it easier for researchers and developers
to exchange and collaborate on software, as dependencies and setup are kept to a
minimum.

Similar benefits hold in the case of specialized languages
for numeric and statistical programming
such as \textsc{matlab} and R \citep{r}.
In comparison to these, though, Python has the distinct advantage
that it is a \textit{general purpose} language,
while NumPy and SciPy extend it with functionality
similar to that offered by \textsc{matlab} and R.
Python has strong language and standard library support for such tasks as
string/text processing, interprocess communication, networking
and many of the other auxiliary tasks that machine learning programs
(whether academic or commercial) routinely need to perform.
While support for many of these tasks is improving in languages such as
\textsc{matlab} and R, they still lag behind Python in their general purpose
applicability.
In many applications of machine learning these tasks, such as data access,
data preprocessing and reporting, can be a more significant task than applying
the actual learning algorithm.

Within the realm of Python,
a package that deserves mention is the Gensim topic modeling toolkit
\citep{rehurek2010gensim},
which exemplifies a different style of API design
geared toward scalable processing of ``big data''.
Gensim's method of dealing with large datasets is to use algorithms
that have $O(1)$ space complexity and can be updated online.
The API is designed around the Python concept of an \textit{iterable}
(supported in the language by a restricted form of co-routines called
\textit{generators}).
While the text vectorizers part of \sklearn
also use iterables to some extent,
they still produce entire sparse matrices, intended to be used for batch or
mini-batch learning. This is the case
even in the stateless, O(1) memory vectorizers
that implement the hashing trick of \citet{weinberger2009}.
This way of processing, as argued earlier in Section~\ref{sec:arrays},
reduces various forms of overhead
and allows effective use of the vectorized operations provided by NumPy and
SciPy.  We make no attempt to hide this batch-oriented processing from the user,
allowing control over the amount of memory actually dedicated
to \sklearn algorithms.

>>>>

\sklearn is a machine learning library for the Python language.\footnote{\url{http://scikit-learn.org}}
It is designed to be simple and efficient, accessible to
non-experts, and reusable in various contexts. In this article, we present and
discuss the uniform application programming interface shared
by all learning and processing units in the library,
and the way it facilitates composition and reusability. We also comment on
implementation details specific to the Python ecosystem
and analyze potential obstacles faced by users and developers of the library.

@@

The \sklearn project \citep{pedregosa2011} provides an open source machine learning
library for the Python programming language. The ambition of the project is to
provide efficient and well-established machine learning tools within a
programming environment that is accessible to non-machine learning experts and
reusable in various scientific areas. The project is not a novel domain-specific
language, but a library that provides machine learning idioms to a
general-purpose high-level language. Among other things, it includes classical learning
algorithms, model evaluation and selection tools, as well as
preprocessing procedures. The library is
distributed under the simplified BSD license, encouraging its use in both
academic and commercial settings.

@@

In \sklearn, the data representation is kept as close as possible
to the matrix representation: datasets are encoded as
two-dimensional NumPy arrays or SciPy sparse matrices.
While these may seem rather unsophisticated data representations when
compared to more object-oriented constructs, such as the ones used by
Weka \citep{hall2009weka}, they bring the prime advantage of allowing us to rely
on efficient NumPy and SciPy vectorized operations while keeping
the code short and readable.  This design choice has also been motivated by
the fact that, given their pervasiveness in many other scientific Python
packages, many scientific users of Python are already familiar with NumPy dense
arrays and SciPy sparse matrices.
From a practical point of view, these formats also provide a collection of
data loading and conversion tools which make them very easy to use in many
contexts. Moreover, for tasks where the inputs are text files or semi-structured
objects, we provide \textit{vectorizer} objects that efficiently convert such
data to the NumPy or SciPy formats.

The public interface is oriented towards processing a batch of samples,
rather than a single sample, in each API call. While classification
and regression algorithms can make predictions for single samples,
\sklearn objects are not optimized for this use case. (The online learning
algorithms in the library are intended to take mini-batches.) Batch processing makes
optimal use of NumPy and SciPy by preventing the overhead inherent to Python
function calls or due to per-element dynamic type checking. Although this might
seem to be an artifact of the Python language, and therefore an implementation
detail that leaks into the API, we argue that APIs should be designed so as not
to tie a library to a suboptimal implementation strategy.
Batch processing also enables fast implementations in lower-level languages, where memory
hierarchy effects and the possibility of internal parallelization come into
play.

@@

From the start, the choice to let a single object serve dual purpose as
estimator and model has mostly been driven by usability and technical
considerations. From the user point of view, having two coupled instances (i.e.,
an estimator object, used as a factory, and a model object, produced by the
estimator) decreases the ease of use and is also more likely to confuse newcomers.
From the developer point of view, decoupling
estimators from models also creates parallel class hierarchies and increases the
overall maintenance complexity of the project. For these practical reasons, we
believe that decoupling estimators from models is not worth the effort. A good
reason for decoupling however, would be that it makes it possible to ship a
model in a new environment without having to deal with potentially complex
software dependencies. Such a feature could however still be implemented in
\sklearn by making estimators able to export a fitted model, using the
information from its public attributes, to an agnostic model description such as
PMML~\citep{guazzelli2009pmml}.

@@

scaler = StandardScaler()
scaler.fit(X_train)
X_train = scaler.transform(X_train)
\end{pythoncode}
Of course, in practice, it is important to apply the same preprocessing to the
test data \texttt{X\_test}. Since a \texttt{StandardScaler} estimator stores the
mean and standard deviation that it computed for the training set, transforming
an unseen test set \texttt{X\_test} maps it into the appropriate region of
feature space:
\begin{pythoncode}
X_test = scaler.transform(X_test)
\end{pythoncode}
Transformers also include a variety of learning algorithms, such as
dimensionality reduction (PCA, manifold learning), kernel approximation,
and other mappings from one feature space to another.

Every transformer allows \texttt{fit(X).transform(X)}
to be written as \texttt{fit\_transform(X)}.
The combined \texttt{fit\_transform} method prevents repeated computations.
Depending on the transformer,
it may skip only an input validation step,
or in fact use a more efficient algorithm for the transformation.
In the same spirit, clustering algorithms provide a
\texttt{fit\_predict} method
that is equivalent to \texttt{fit} followed by \texttt{predict},
returning cluster labels assigned to the training samples.

@@

Optionally, the model selection algorithms
also take a cross-validation scheme and a score function.  \sklearn provides
various such schemes, including $k$-fold (default),
stratified $k$-fold and leave-one-out.
The score function used by default is the estimator's \texttt{score} method,
but the library provides a variety of
alternatives that the user can choose from,
including accuracy, AUC and $F_1$ score for classification,
$R^2$ score and mean squared error for regression.

@@

In other places of the library, in particular in the vectorization code
for unstructured input, the toolkit is also designed to be
extensible. Here, estimators provide hooks for user-defined code: objects or
functions that follow a specific API can be given as arguments at vectorizer
construction time. The library then calls into this code,
communicating with it by passing objects of standard Python/NumPy types,
and keeping it agnostic of the \sklearn class hierarchy.

Our rule of thumb is that user code should not be tied to \sklearn---which is a
\textit{library}, and not a framework. This principle avoids a
common problem with object-oriented design, which is that users wanting a
``banana'' should not get ``a gorilla holding the banana and the entire jungle''
(J.~Armstrong, cited by \citealp[p.~213]{seibel2009coders}).
Programs using \sklearn should not be intimately tied to it,
so that their code can be reused with other toolkits or in other contexts.

@@

Our implementation guidelines emphasize writing efficient but readable code.
In particular, we focus on making the codebase maintainable and
understandable in order to favor external contributions. Whenever practical,
algorithms implemented in \sklearn are written in Python,
using NumPy vector operations for numerical work.
This allows the code to remain concise, readable and efficient.
For critical algorithms that cannot be easily and efficiently
expressed as NumPy operations, we rely on Cython \citep{behnel2011cython}
to achieve competitive performance and scalability. Cython is a
compiled programming language that extends Python with static typing. It
produces efficient C extension modules that are importable from the
Python run-time system. Examples of Cython code in \sklearn are
stochastic gradient descent for linear models, some clustering
algorithms, and decision trees.

To facilitate the installation of \sklearn,
the set of external dependencies is kept to a bare minimum:
only Python, NumPy and SciPy are required for a functioning installation.
Binary distributions of these are available for the major platforms.
Visualization functionality uses on Matplotlib \citep{hunter2007matplotlib}
and Graphviz \citep{gansner2000},
but neither is required to perform machine learning or prediction.
When feasible, external libraries are integrated into the codebase.
In particular, \sklearn includes modified versions of \textsf{LIBSVM} and \textsf{LIBLINEAR}
\citep{chang2011libsvm,fan2008}, both written in C{}\verb!++!
and wrapped using Cython.

@@

\sklearn competes with a many machine learning packages,
both open source and proprietary.
Some of those, including Weka~\citep{hall2009weka} or
Orange~\citep{Demsar2004}, offer APIs but actually focus on the use of a graphical user interface (GUI)
which allows novices to easily apply machine learning algorithms. By
contrast, the target audience of \sklearn is capable of programming, and
therefore we focus on developing a usable and consistent API, rather than expend
effort into creating a GUI\@. In addition, while GUIs are useful tools, they
sometimes make reproducibility difficult in the case of complex workflows
(although those packages usually have developed a GUI for managing complex
tasks).

Other existing machine learning packages
such as SofiaML\footnote{\url{https://code.google.com/p/sofia-ml}}
and Vowpal~Wabbit\footnote{\url{http://hunch.net/~vw}}
are intended to be used as command-line tools
(and sometimes do not offer any kind of API).
While these packages have the advantage
that their users are not tied to a particular programming language,
users will find that they still need programming to process input/output,
and will do so in a variety of languages.
By contrast, \sklearn allows users to implement that entire workflow
in a single program, written in a single language,
and developed in a single working environment.
This also makes it easier for researchers and developers
to exchange and collaborate on software, as dependencies and setup are kept to a
minimum.

Similar benefits hold in the case of specialized languages
for numeric and statistical programming
such as \textsc{matlab} and R \citep{r}.
In comparison to these, though, Python has the distinct advantage
that it is a \textit{general purpose} language,
while NumPy and SciPy extend it with functionality
similar to that offered by \textsc{matlab} and R.
Python has strong language and standard library support for such tasks as
string/text processing, interprocess communication, networking
and many of the other auxiliary tasks that machine learning programs
(whether academic or commercial) routinely need to perform.
While support for many of these tasks is improving in languages such as
\textsc{matlab} and R, they still lag behind Python in their general purpose
applicability.
In many applications of machine learning, these tasks, such as data access,
data preprocessing and reporting, can take more work than applying
the actual learning algorithm.

Within the realm of Python,
a package that deserves mention is the Gensim topic modeling toolkit
\citep{rehurek2010gensim},
which exemplifies a different style of API design
geared toward scalable processing of ``big data''.
Gensim's method of dealing with large datasets is to use algorithms
that have small space complexity and can be updated online.
The API is designed around the Python concept of an ``iterable''
(supported in the language by a restricted form of co-routines called generators).
While the text vectorizers in \sklearn
also leverage iterables to some extent,
they still produce entire sparse matrices,
intended to be used for batch or mini-batch learning.
This way of processing, as argued earlier in Section~\ref{sec:arrays},
reduces various forms of overhead
and allows effective use of the vectorized operations provided by NumPy and
SciPy.  We make no attempt to hide this batch-oriented processing from the user,
allowing control over the amount of memory actually dedicated
to \sklearn algorithms.
